---
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
nsims <- 100000 #set number of simulations
library(mvtnorm, quietly = TRUE)
library(MASS, quietly = TRUE)
library(afex, quietly = TRUE)
library(emmeans, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(gridExtra, quietly = TRUE)
library(reshape2, quietly = TRUE)
library(pwr, quietly = TRUE)

# Install functions from GitHub by running the code below:
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/ANOVA_design.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/ANOVA_power.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/helper_functions/power_oneway_between.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/helper_functions/power_oneway_within.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/helper_functions/power_twoway_between.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/helper_functions/power_2x2_within.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/helper_functions/plot_power_oneway_between.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/helper_functions/plot_power_oneway_within.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/helper_functions/plot_power_twoway_between.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/helper_functions/plot_power_2x2_within.R")
```

## Power for Three-way Interactions

There are almost no software solutions that allow researchers to perform power anaysis for more complex designs. Through simulation, it is relatively straightforward to examine the power for designs with multiple factors with many levels. 

Let's start with a 2x2x2 between subjects design. We collect 50 participants in each between participant condition (so 400 participants in total - 50x2x2x2). 

```{r}
string <- "2b*2b*2b"
n <- 50
mu <- c(1, 1, 6, 1, 6, 6, 1, 6) 
sd <- 10
r <- 0.0
labelnames <- c("Size", "big", "small", "Color", "green", "red", 
                "CognitiveLoad", "present", "absent") #

design_result <- ANOVA_design(string = string,
                   n = n, 
                   mu = mu, 
                   sd = sd, 
                   labelnames = labelnames)

# Power for the given N in the design_result
ANOVA_power(design_result, nsims = nsims)

```

A Three-Way ANOVA builds on the same principles as a One_Way ANOVA. We look at whether the differences between groups are large, compared to the standard deviation. For the main effects we simply have 2 groups of 200 participants, and 2 means. If the population standard deviations are identical across groups, this is not in any way different from a One-Way ANOVA. Indeed, we can show this by simulating a One-Way ANOVA, where instead of 8 conditions, we have two conditions, and we average over the 4 groups of the other two factors. For example, for the main effect of size above:

```{r}
string <- "2b"
n <- 200
mu <- c(mean(c(1, 1, 6, 1)), mean(c(6, 6, 1, 6)))
sd <- 10
labelnames <- c("Size", "big", "small")

design_result <- ANOVA_design(string = string,
                   n = n, 
                   mu = mu, 
                   sd = sd, 
                   labelnames = labelnames)

# Power based on simulations
ANOVA_power(design_result, nsims = nsims)

# Power based on analytical solution
power_oneway_between(design_result)$power #using default alpha level of .05

```

Similarly, we can create a 2 factor design where we average over the third factor, and recreate the power analysis for the Two-Way interaction. For example, we can group over the Cognitive Load condition, and look at the Size by Color Interaction:

```{r}
string <- "2b*2b"
n <- 100
mu <- c(mean(c(1, 1)), mean(c(6, 1)), mean(c(6, 6)), mean(c(1, 6)))
sd <- 10
labelnames <- c("Size", "big", "small", "Color", "green", "red")

design_result <- ANOVA_design(string = string,
                   n = n, 
                   mu = mu, 
                   sd = sd, 
                   labelnames = labelnames)

# Power based on simulations
ANOVA_power(design_result, nsims = nsims)

# Power based on analytical solution
power_res <- power_twoway_between(design_result) #using default alpha level of .05

power_res$power_A
power_res$power_B
power_res$power_AB

```

