---
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(mvtnorm)
library(afex)
library(emmeans)
library(ggplot2)
library(gridExtra)
library(reshape2)
library(pwr)

# Install functions from GitHub by running the code below:
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/ANOVA_design.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/ANOVA_power.R")

```

## Setting the correlation matrix

In a 2x2 design, with factors A and B, each with 2 levels, there are 6 possible comparisons that can be made.

1. A1 vs. A2
2. A1 vs. B1
3. A1 vs. B2
4. A2 vs. B1
5. A2 vs. B2
6. B1 vs. B2

The number of possible comparisons is the product of the levels of all factors squared minus the product of all factors, divided by two. For a 2x2 design where each factor has two levels, this is:

```{r}
(((2*2)^2)-(2*2))/2
```

The number of possible comparisons increases rapidly when adding factors and levels for each factor. For example, for a 2x2x4 design it is:

```{r}
(((2*2*4)^2)-(2*2*4))/2
```

Each of these comparisons can have their own correlation if the factor is manipulated within subjects (if the factor is manipulated between subjects the correlation is 0). These correlations determine the covariance matrix. Potvin and Schutz (2000) surveyed statistical tools for power analysis and conclude that most software packages are limited to one factor repeated measure designs and do not provide power calculations for within designs with multiple factor (which is still true for software such as G*Power). Furthermore, software solutions which were available at the time (DATASIM by Bradley, Russel, & Reeve, 1996) required researchers to assume correlations were of the same magnitude for all within factors, which is not always realistic. If you do not want to assume equal correlations for all paired comparisons, you can specify the correlation for each possible comparison.  

The order in which the correlations are entered in the vector should match the covariance matrix.
The order for a 2x2 design is given in the 6 item list above. The general pattern is that the matrix is filled from top to bottom, and left to right, illustrated by the increasing correlations in the table below.

             a1_b1           a1_b2           a2_b1           a2_b2
a1_b1        1.00            0.91            0.92            0.93
a1_b2        0.91            1.00            0.94            0.95
a2_b1        0.92            0.94            1.00            0.96
a2_b2        0.93            0.95            0.9             1.00

The diagonal is generated dynamically (based on the standard deviation).

We would enter this correlation matrix as:

```{r}
design_result <- ANOVA_design(string = "2w*2w",
                              n = 80,
                              mu = c(1.1, 1.2, 1.3, 1.4),
                              sd = 1,
                              r <- c(0.91, 0.92, 0.93, 0.94, 0.95, 0.96),
                              p_adjust = "none",
                              labelnames = c("color", "red", "blue", "speed", "fast", "slow"))

```

We can check the correlation matrix by asking for it from the design_result object to check if it was entered the way we wanted:

```{r}
design_result$cor_mat
```

