---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
nsims <- 10000 #set number of simulations
```

## Validation for One-Way ANOVA

Using the formulas below, we can calculate the means for between designs with one factor (One-Way ANOVA). Using the formula also used in Albers & Lakens (2018), we can determine the means that should yield a specified effect sizes (expressed in Cohen's f):

```{r}
mu_from_ES <- function(K, ES){ # provides the vector of population means for a given population ES and nr of groups
  f2 <- ES/(1-ES)
  if(K == 2){
    a <- sqrt(f2)
    muvec <- c(-a,a)
  }
  if(K == 3){
    a <- sqrt(3*f2/2)
    muvec <- c(-a, 0, a)
  }
  if(K == 4){
    a <- sqrt(f2)
    muvec <- c(-a, -a, a, a)
  } # note: function gives error when K not 2,3,4. But we don't need other K.
  return(muvec)
}

```

Eta-squared (idential to partial eta-squared for One-Way ANOVA's) has benchmarks of .0099, .0588, and .1379 for small, medium, and large effect sizes (cohen, 1988). 

## Installation

We install the functions:

```{r}
# Install the two functions from GitHub by running the code below:

source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/ANOVA_design.R")
source("https://raw.githubusercontent.com/Lakens/ANOVA_power_simulation/master/ANOVA_power.R")
```

## Three conditions, small effect size

We can simulate a one-factor anova setting means to achieve a certain effect size. 

```{r}
K <- 3
ES <- .0099
mu <- mu_from_ES(K = K, ES = ES)
n <- 5000
sd <- 1
r <- 0
string = paste(K,"b",sep="")
```



```{r, message=FALSE, warning=FALSE}
design_result <- ANOVA_design(string = string,
                   n = n, 
                   mu = mu, 
                   sd = sd, 
                   r = r, 
                   p_adjust = "none")

ANOVA_power(design_result, nsims = nsims)

```

Results according to pwr package for the ANOVA

```{r}
library(pwr)
f2 <- ES/(1-ES)
f <- sqrt(f2)
pwr.anova.test(f=f,k=K,n=n,sig.level=0.05)
```

## Four conditions, medium effect size

We can simulate a one-factor anova setting means to achieve a certain effect size. 

```{r}
K <- 4
ES <- .0588
mu <- mu_from_ES(K = K, ES = ES)
n <- 5000
sd <- 1
r <- 0
string = paste(K,"b",sep="")
```



```{r, message=FALSE, warning=FALSE}
design_result <- ANOVA_design(string = string,
                   n = n, 
                   mu = mu, 
                   sd = sd, 
                   r = r, 
                   p_adjust = "none")

ANOVA_power(design_result, nsims = nsims)

```

Results according to pwr package for the ANOVA

```{r}
library(pwr)
f2 <- ES/(1-ES)
f <- sqrt(f2)
pwr.anova.test(f=f,k=K,n=n,sig.level=0.05)
```


## Two conditions, large effect size

We can simulate a one-factor anova setting means to achieve a certain effect size. 

```{r}
K <- 2
ES <- .1379
mu <- mu_from_ES(K = K, ES = ES)
n <- 5000
sd <- 1
r <- 0
string = paste(K,"b",sep="")
```



```{r, message=FALSE, warning=FALSE}
design_result <- ANOVA_design(string = string,
                   n = n, 
                   mu = mu, 
                   sd = sd, 
                   r = r, 
                   p_adjust = "none")

ANOVA_power(design_result, nsims = nsims)

```

Results according to pwr package for the ANOVA

```{r}
library(pwr)
f2 <- ES/(1-ES)
f <- sqrt(f2)
pwr.anova.test(f=f,k=K,n=n,sig.level=0.05)
```

