---
title             : "Simulation-Based Power-Analysis for Factorial ANOVA Designs"
shorttitle        : "ANOVA Power"

author: 
  - name          : "Dani&euml;l Lakens"
    affiliation   : "1"
    corresponding : yes
    address       : "ATLAS 9.402, 5600 MB, Eindhoven, The Netherlands"
    email         : "D.Lakens@tue.nl"
  - name          : "Aaron R. Caldwell"
    affiliation   : "2"

affiliation:
  - id            : "1"
    institution   : "Eindhoven University of Technology, The Netherlands"
  - id            : "2"
    institution   : "Department of Health, Human Performance and Recreation, University of Arkansas, USA"

author_note: |
  All code used to create this manuscript is provided in an OSF repository at https://osf.io/xxxxx/.

abstract: |
  Researchers often use an analysis of variance (ANOVA) when reporting results of study. In order for an ANOVA to be informative researchers need to ensure their study is adequately powered. Yet, power analyses for factorial ANOVA designs are often challenging. First, current software solutions do not enable power analyses for complex designs with many within-subject factors. Second, power analyses often need partial eta-squared or Cohen's f as input, but these effect sizes do not generalize to different experimental designs. We have created R functions and an online Shiny app that performs simulations for ANOVA designs. Our functions allow for up to three within- or between-subject factors, with an unlimited number of levels. Predicted effects are entered by specifying means, standard deviations, and correlations (for within-subject factors). The simulation provides power calculations for all ANOVA main effects and interactions. In addition, power calculations are provided for pairwise comparisons and there are range of options to correct for multiple comparisons.  The simulation plots *p*-value distributions for all tests.  This tutorial will demonstrate how to perform power analysis for ANOVA designs. The simulations illustrate important factors that determine the statistical power of factorial ANOVA designs. The code and Shiny app will enable researchers without extensive programming experience to perform power analyses for a wide range of ANOVA designs.
  
keywords          : "power analysis, ANOVA, hypothesis test, sample size justification, repeated measures, mixed designs"
wordcount         : Main text contains `r wordcountaddin::word_count("0.1_Simulation_Based_Power_Analysis_For_Factorial_ANOVA_Designs.Rmd")` words.


bibliography      : ["anova_power.bib"]

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : no

class             : "jou, draftfirst, a4paper"
output            : papaja::apa6_pdf
---
```{r load_packages, include=FALSE}
#Set to low number when test-compiling, set to 100.000 for final manuscript.
nsims <- 10000

#Set manuscript seed

manuscript_seed = 2019

library("papaja")
require(MASS, quietly = TRUE)
require(ggplot2, quietly = TRUE)
require(gridExtra, quietly = TRUE)
require(reshape2, quietly = TRUE)
require(RColorBrewer, quietly = TRUE)

if(!require(ANOVApower)){devtools::install_github("Lakens/ANOVApower")}
library(ANOVApower)

wordcountaddin::word_count("0.1_Simulation_Based_Power_Analysis_For_Factorial_ANOVA_Designs.Rmd")

```

When a researcher aims to test hypotheses based on an analysis of variance (ANOVA), the sample size of the study should be justified based on the statistical power of the test. 
Statistical power is the probability of rejecting the null-hypothesis, given a specified effect size, alpha level, and sample size. 
There is a high  probability declaring there is no effect when there is low power; even when the effect of interest may exist.
Several excellent resources exist that explain power analyses, including books [@aberson_applied_2019; @cohen_statistical_1988], general reviews [@maxwell_sample_2008], and practical primers [@brysbaert_how_2019; @perugini_practical_2018]. 
Whereas power analyses for simple comparisons are relatively easy to perform, power analysis for factorial ANOVA designs is a challenge. 
Available software solutions do not provide easy options for more complex designs (e.g., a 2x2x2 design, where the first factor is manipulated between participants, and the last two factors are manipulated within participants). 
The predicted effects often need to be specified as Cohen's f or partial eta squared ($\eta_p^2$), which are not the most intuitive way to specify a hypothesized pattern of results, and do not generalize to different experimental designs.
Simulations based on a specified pattern of means and covariance matrix (i.e., specific standard deviation and correlation) provides a more flexible approach to power analyses. However, such simulations traditionally have required extensive programming knowledge.

In this manuscript we provide R functions and a Shiny app that can be used to perform power analyses for factorial ANOVA designs based on simulations. Both the R functions and Shiny app calculate the statistical power based on a predicted pattern of means, standard deviations, and correlations (for within-subject factors). 
By simulating different factorial designs, researchers can gain a better understanding of the factors that determine the statistical power of hypothesis tests. 
After providing an introduction to statistical power in general, and for the *F*-test specifically, we will demonstrate how the power of factorial ANOVA designs depend on the pattern of means across conditions, the number of factors and levels, and the sample size. 
We also illustrate the importance of controlling the Type 1 error rate in exploratory ANOVA's to maintain the desired alpha level when performing multiple comparisons.

# Calculating Power in ANOVA Designs

Imagine you plan to perform a study in which participants interact with an artificial voice assistant who sounds either cheerful or sad.
You measure how much 80 participants in each condition enjoy to interact with the voice assistant on a line marking scale (coded continuously from -5 to 5) and observe a mean of 0 in the sad condition, and a means of 1 in the cheerful condition, with an estimated standard deviation of 2.
After submitting your manuscript for publication, reviewers ask you to add a study with a neutral control condition to examine whether cheerful voices increase, or sad voices decrease enjoyment (or both). 
Depending on what the mean enjoyment in the neutral condition is, which sample size would you need to have a high powered study for the expected pattern or means? 
A collaborator suggests to switch from a between-subject design to a within-subject design, to more efficiently collect the data.
What is the consequence of switching from a between-subject to a within-subject design on the required sample size?
The effect size in the first study could be considered 'medium' based on the benchmarks by @cohen_statistical_1988, but does it make sense to plan for a 'medium' effect size in either the between-subject or within-subject ANOVA design?
And if you justify the sample size based on the power for the main effect for the ANOVA, will the study also have sufficient statistical power for the independent comparisons between conditions (or vice versa)?

```{r d-plot, fig.width=7, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE, results='hide', fig.cap="Distribution of Cohen's d under the null-hypothesis (grey curve) and alternative hypothesis assuming d = 0.5 (black curve)."}
    N <- 80
    d <- 0.5
    p_upper<-.05
    p_lower<-0
    ymax<-25 #Maximum value y-scale (only for p-curve)
    
    ncp<-(d*sqrt(N/2)) #Calculate non-centrality parameter d
    low_x<--1
    high_x<-1.5

    #calc d-distribution
    x=seq(low_x,high_x,length=10000) #create x values
    d_dist<-dt(x*sqrt(N/2),df=(N*2)-2, ncp = ncp)*sqrt(N/2) #calculate distribution of d based on t-distribution
    #Set max Y for graph
    y_max<-max(d_dist)+1
    #create plot
    par(bg = "white")
    plot(-10,xlim=c(low_x,high_x), ylim=c(0,y_max), xlab=substitute(paste("Cohen's ", delta)), ylab="Density",main="")
    axis(side = 1, at = seq(low_x,high_x,0.1), labels = FALSE)
    lines(x,d_dist,col='black',type='l', lwd=2)
    #add d = 0 line
    d_dist<-dt(x*sqrt(N/2),df=(N*2)-2, ncp = 0)*sqrt(N/2)
    lines(x,d_dist,col='grey',type='l', lwd=2)
    #Add Type 1 error rate right
    crit_d<-abs(qt(p_upper/2, (N*2)-2))/sqrt(N/2)
    y=seq(crit_d,10,length=10000) 
    z<-(dt(y*sqrt(N/2),df=(N*2)-2)*sqrt(N/2)) #determine upperbounds polygon
    polygon(c(crit_d,y,10),c(0,z,0),col="lightgrey")
    #Add Type 1 error rate left
    crit_d<--abs(qt(p_upper/2, (N*2)-2))/sqrt(N/2)
    y=seq(-10, crit_d, length=10000) 
    z<-(dt(y*sqrt(N/2),df=(N*2)-2)*sqrt(N/2)) #determine upperbounds polygon
    polygon(c(y,crit_d,crit_d),c(z,0,0),col="lightgrey")
    #Add Type 2 error rate
    crit_d<-abs(qt(p_upper/2, (N*2)-2))/sqrt(N/2)
    y=seq(-10,crit_d,length=10000) 
    z<-(dt(y*sqrt(N/2),df=(N*2)-2, ncp=ncp)*sqrt(N/2)) #determine upperbounds polygon
    polygon(c(y,crit_d,crit_d),c(0,z,0),col="darkgrey")
    segments(crit_d, 0, crit_d, y_max-0.8, col= 'black', lwd=2, lty = "dashed")
    segments(-crit_d, 0, -crit_d, y_max-0.8, col= 'black', lwd=2, lty = "dashed")
```

```{r eta-plot, fig.width=7, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE, results='hide', fig.cap="Distribution of eta-squared under the null-hypothesis (grey curve) and alternative hypothesis assuming partial eta-squared = 0.0588 (black curve)."}

    N<-80
    J<-2
    eta_pop<-0.0588
    alpha<-.05
    xmax<-.2
    ymax<-.6
    
    xmin <- 0
    #Calculations
    df1 <- J-1
    df2 <- J*(N - 1)
    Ntot = N*J
    ncp = Ntot/(1/eta_pop-1)
    crit_f <- qf(1 - alpha, df1, df2)
    
    x=seq(xmin,xmax,length=1000)
    #F-value function
    eta_pop_dist <- function(x) df((x*df2)/(df1-x*df1), df1, df2, ncp)
    par(bg = "white")
    plot(-10,xlab=substitute(paste(eta[p]^2)), ylab="Density", axes=FALSE,
         main = "", xlim=c(0,xmax),  ylim=c(0, ymax))
    #title in shiny app: main=substitute(paste("distribution for ",eta[p]^2 == eta_pop," for ",J," groups and ", N, " observations per group."
    axis(side=1, at=seq(0,xmax, 0.02), labels=seq(0,xmax, 0.02))
    axis(side=2)
    ncp<-0 #set ncp to 0 to plot the F distribution for null effect.
    eta_pop_crit<-(crit_f*df1)/(crit_f*df1+df2)
    
    curve(eta_pop_dist, 0.00000000001, 1, n=10000, col="grey", lwd=2, add=TRUE)
    x=seq(eta_pop_crit,xmax,length=10000) 
    z<-df((x*df2)/(df1-x*df1), df1, df2) #determine upperbounds polygon
    polygon(c(eta_pop_crit,x,1),c(0,z,0),col="lightgray") #draw polygon
    #Add Type 2 error rate
    ncp = Ntot/(1/eta_pop-1)
    curve(eta_pop_dist, 0.00000000001, 1, n=10000, col="black", lwd=2, add=TRUE)
    y=seq(0.00000000001,eta_pop_crit,length=10000) 
    z<-df((y*df2)/(df1-y*df1), df1, df2, ncp) #determine upperbounds polygon
    polygon(c(y,eta_pop_crit,eta_pop_crit),c(0,z,0),col="darkgray")
    segments(eta_pop_crit, 0, eta_pop_crit, ymax-0.03, col= 'black', lwd=2, lty = "dashed")
    
```

Let's consider the initial study described above, where enjoyment is measured when interacting with a cheerful or sad voice assistant.
We can test the difference between two means with a *t*-test or a one-way ANOVA, and the two tests are mathematically equivalent. 
Figure \ref{fig:d-plot} and Figure \ref{fig:eta-plot} visualize the distribution of the effect sizes Cohen's d (for the *t*-test) and $\eta_p^2$ (for the *F*-test) that should be observed when there is no effect (grey curves) and when the observed difference between means equals the true effect (black curves)^[For Shiny apps that allow you to dynamically change these figures, go to http://shiny.ieis.tue.nl/d_p_power/ and http://shiny.ieis.tue.nl/f_p_power/].
In both figures the light grey areas under the null-distribution mark the observed results that would lead to a Type 1 error (observing a statistically significant result if the null-hypothesis is true) and the dark grey areas under the curve marks the observed effect sizes that would lead to a Type 2 error.
A statistically significant result is observed whenever the observed effect size is larger than the critical value.
Critical values are often expressed as *t*-values or *F*-values, but for a given sample size can be expressed as effect sizes.
Given the sample size of 80 participants per group, effects are statistically significant when they are larger than d = `r crit_d` in a *t*-test, or $\eta_p^2$ = `r eta_pop_crit` for the *F*-test.
The goal of a power analysis is to choose a sample size so that the probability of observing a statistically significant effect reaches a desired probability. 
To calculate power, one has to specify the alternative hypothesis (the black curves in Figure \ref{fig:d-plot} and \ref{fig:eta-plot}).
If we assume the true effect size is d = 0.5 or $\eta_p^2$ = 0.0588, and data is collected from 80 participants in each condition, in the long run 88.16% of the observed effects will yield statistically significant results. 

Two relationships between the *t*-test and the *F*-test when comparing two means are worth pointing out. 
The *t*-test examines the difference between means $(m_1 - m_2)$, and the *F*-test computes the ratio of the between group variance and the within group variance. 
For two groups, the variance of the difference between two means is $(m_1 - m_2)^2$, and therefore $F = t^2$.
The main difference between the *t*-test and the *F*-test is that the *F*-test generalizes to more than two groups.
Second, for two groups the effect size for an ANOVA, Cohen's f, is half the size of the effect size for standardized mean differences, Cohen's d, or $f = \frac{1}{2}d$. 
Cohen's d is calculated by dividing the difference between means by the standard deviation, or 
\begin{equation}
d = \frac{m_1-m_2}{\sigma}.
\end{equation}
If we have two groups with means of 1 and 2, and the standard deviation is 2, Cohen's d is (2-1)/2, or 0.5.
Cohen's f is the standard deviation of the population means divided by the population standard deviation [@cohen_statistical_1988], or: 
\begin{equation}
f = \frac{\sigma _{ m }}{\sigma}
\end{equation}
where for equal sample sizes,
\begin{equation}
\sigma _{ m } = \sqrt { \frac { \sum_ { i = 1 } ^ { k } ( m _ { i } - m ) ^ { 2 } } { k } }.
\end{equation}
Because Cohen's f is the effect size that is an essential factor when computing the power for factorial ANOVA designs, it is worth illustrating how it is calculated in an example.
If we again take two means of 1 and 2, and a standard deviation of 2, the grand mean is 1.5. 
We subtract each condition mean from the grand mean, take the square, calculate the sum of squares, divide it by two, and take the square root. 
$\sigma_m = \sqrt{\frac{(1-1.5)^2+(2-1.5)^2}{2}} = \sqrt{\frac{0.25+0.25}{2}} = 0.5$, and $f = \frac{0.5}{2} = 0.25.$
We confirm that for a two group comparison, Cohen's f is half as large as Cohen's d.
Popular power analysis software such as G\*Power [@faul_gpower_2007] also allows researchers to enter the effect size for the power analysis as partial eta-squared ($\eta_p^2$).
Partial eta-squared can be converted into Cohen's f: 
\begin{equation}
f = \sqrt{\frac{\eta_p^2}{1-\eta_p^2}} \label{eq:eta-to-f}
\end{equation}
and Cohen's f can be converted into partial eta-squared:
\begin{equation}
\eta_p^2 = \sqrt{\frac{f^2}{f^2+1}} \label{eq:f-to-eta}
\end{equation}
In the example above, $\eta_p^2 = 0.25^2/(0.25^2+1) = 0.0588$. 
To calculate the statistical power assuming a specific true effect size we need the noncentrality parameter of the distribution. 
In both Figure \ref{fig:d-plot} and Figure \ref{fig:eta-plot} we see examples of the non-central *t*-distribution and non-central *F*-distribution, or the shape of the expected test statistics when there is a true effect (the black curves).
Power calculations rely on the noncentrality parameter often referred to as lambda ($\lambda$). In a between-participants one-way ANOVA lambda is calculated as:
\begin{equation}
\lambda = f^2 \times N \label{eq:lambda}
\end{equation}
where f is Cohen's f and N is the total sample size. 
Based on $\lambda$ (which specifies the shape of the expected distribution under the specified alternative hypothesis) and the critical test statistic (which specifies the part of the distribution that is more extreme than the test statistic needed for a statistically significant test result) we can calculate how much of the distribution under the alternative hypothesis will be statistically significant in the long run (i.e., the area under the black curve in Figure \ref{fig:d-plot} and \ref{fig:eta-plot} to the right of the critical effect size). 

# Simulating Statistical Power for Different Factorial Designs

We have created R code and a Shiny app that simulate factorial ANOVA designs. 
At the core, the code generates data for each condition in the design, performs an ANOVA. 
The simulation also performs *t*-tests for all independent comparisons between conditions.
Then the percentage of significant results (i.e., the power) is calculated (after correcting the alpha level for multiple comparisons if desired) and average effect size estimates are computed for all tests.
The software requires specifying the design of the study, by providing the number of levels for each factor, and indicating whether the factor is measured between or within participants.  Our initial study above is a '2b' or two level between-participant design.
For ease of interpretation the factors and levels can be named (for our example, factor = "Condition" with levels = "cheerful" and "sad").
The planned sample size should be specified, which is 80 participants in each between-participant condition.
The means for each condition should be specified (0 and 1), as well as the standard deviation (2). 
For within-subject designs the correlation between variables should be specified.
When these variables are provided, the design is set up for the simulation.
For a visual confirmation of the input, a figure is created that displays the means and standard deviation (see Figure \ref{fig:mean-plot2}).

```{r mean-plot2, fig.width=7, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE, results='hide', fig.cap="Vizualization for the expected means and standard deviations for the two conditions."}

design_result <- ANOVA_design(string = "2b", n = 80, mu = c(1, 0), sd = 2, labelnames = c("condition", "cheerful", "sad"))

power_result <- ANOVA_power(design_result, alpha_level = 0.05, p_adjust = "none", nsims = nsims, seed = manuscript_seed, verbose = FALSE)
#Analytic solution
power_result_2 <- power_oneway_between(design_result)$power
```

The results of a simulation study will vary each time the simulation is performed (but can be made reproducible by specifying a 'seed' number).
Results become more stable, the more simulations are performed, but the simulations also take longer. 
To perform the simulations, you specify the number of simulations, the alpha level for the tests, and any adjustment for multiple comparisons for the ANOVA. 
If 100.000 simulations are performed in R with a seed set to 2019 (these settings will be used for all simulation results in this manuscript), we see the statistical power (based on the percentage of *p* < $\alpha$ results) is `r power_result$main_results$power` and the average effect size is `r power_result$main_results$effect_size`.
The simulation also provides the results for the independent comparisons, based on *t*-tests comparing each condition.
Since there are only two groups in this example, the results for the statistical power of the independent comparison is identical, but the effect size, Cohen's d, is `r power_result$pc_results$effect_size`. 

Now that the basic idea behind the simulation is clear, we can start exploring how changing the experimental design influences power. 
Further, answer some of the questions our hypothetical researcher was confronted with when designing a follow-up study. 
We will first examine what happens if we add a third condition to the design.
Let's assume we expect the neutral voice condition to fall either between the cheerful and sad conditions, or perhaps to be equal to the cheerful condition (based on the idea that the sad voice leads to less enjoyment, but the cheerful voice does not lead to more enjoyment).
The design now has 3 between-participant conditions, and we can explore what happens if we would collect 80 participants in each condition.

```{r sim-3, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
design_result_1 <- ANOVA_design(string = "3b", n = 80, mu = c(1, 0.5, 0), sd = 2, labelnames = c("condition", "cheerful", "neutral", "sad"))
power_result_1 <- ANOVA_power(design_result_1, alpha_level = 0.05, p_adjust = "none", nsims = nsims, seed = manuscript_seed, verbose = FALSE)

design_result_2 <- ANOVA_design(string = "3b", n = 80, mu = c(1, 1, 0), sd = 2, labelnames = c("condition", "cheerful", "neutral", "sad"))
power_result_2 <- ANOVA_power(design_result_2, alpha_level = 0.05, p_adjust = "none", nsims = nsims, seed = manuscript_seed, verbose = FALSE)

#analytic solutions
power_oneway_between(design_result_1)$power
power_oneway_between(design_result_1)$Cohen_f
power_oneway_between(design_result_2)$power
power_oneway_between(design_result_2)$Cohen_f

```

If we assume the mean falls exactly between between the cheerful and sad conditions the simulations show the statistical power for our design is reduced to `r power_result_1$main_results$power`%, and the effect size (partial eta-squared) is `r power_result_1$main_results$effect_size`. 
If we assume the mean the mean is equal to the cheerful condition, the power increases to `r power_result_2$main_results$power`%.
Compared to the two group design, three things have changed. 
First, an additional group was added to the design. 
This increases the numerator degrees of freedom, and makes the non-central *F*-distribution more similar to the central *F*-distribution, and reduces the statistical power. 
Second, the total sample size is 50% larger than in the initial two-group study, which increases the statistical power.
Third, the effect size has decreased, which reduces the statistical power.
The exact effect of these three changes on the statistical power is difficult to predict. 
The most important conclusion based on these simulations is that changing an experimental design can have several opposing effects on the power of a study, depending of the pattern of means. 
One cannot assume the effect size remains unchanged, when the design changes.

# Power for Within-Subject Designs

What happens if we would perform the second study as a within-participants design?
Instead of collecting three groups of participants, we only collect one group, and let this group evaluate the cheerful, neutral, and sad voice assistants.
The sample size needed in a within-subject design ($N_W$), relative to the sample needed in between-design ($N_B$), is (from Maxwell & Delaney, 2004, p. 562, formula 47): 
\begin{equation}
N_{W}=\frac{N_{B}(1-\rho)}{a} \label{eq:within-n}
\end{equation}
Here *a* is the number of within-participant levels, $\rho$ is the correlation between the measurements. The formula assumes a normal distribution and compound symmetry, and ignores the difference in degrees of freedom between the two types of tests, so it is a rough (but useful) approximation.
While not always true, the sample size that is needed in a within-participants condition will be smaller than the requires sample size for a between-participants condition when the correlation between dependent measures is positive. Overall, we can assume power will be higher with a stronger correlation between dependent measures.

```{r sim-4, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
design_result_within_1 <- ANOVA_design(string = "3w", n = 80, mu = c(1, 0.5, 0), sd = 2, r = 0.5, labelnames = c("condition", "cheerful", "neutral", "sad"))
power_result_within_1 <- ANOVA_power(design_result_within_1, alpha_level = 0.05, p_adjust = "none", nsims = nsims, seed = manuscript_seed, verbose = FALSE)

#analytic solutions
power_oneway_within(design_result_within_1)$power
```

We can revisit our experiment where we want to collect enjoyment ratings when interacting with with a cheerful, neutral, or sad voice assistant, and explore the power for a within-subject design. 
Assuming means of 1, 0.5, and 0, with a standard deviation of 2, and 80 participants per condition, power for the between-subjects design was `r power_result_1$main_results$power`%.
We need to predict what the correlation between the dependent variables will be, ideally based on data from previous experiments.
It is wise to explore a range of values through a sensitivity analysis when performing the simulations, 
Let's assume our best estimate of the correlation between enjoyment ratings in a within-subject design is r = 0.5.
The power for a repeated-measures ANOVA based on these values, where ratings for the three conditions are collected from 80 participants, is `r power_result_within_1$main_results$power`%.
Power is higher in the within-subject design, even though there are fewer total participants, because all subjects contribute three correlated observations.
Note that simulation studies allow for great flexibility, and the Shiny app and ANOVApower package allow researchers to enter a correlation matrix that specifies the expected correlations between each individual pair of measurements, instead of assuming the correlations between all dependent variables are identical.

Correlated observations have higher power because a positive correlation reduces the standard deviation of the difference scores. 
Because the standardized effect size is the mean difference divided by the standard deviation of the difference scores, the correlation has an effect on the standardized mean difference in a within-subject design, referred to as Cohen's $d_z$ (because it is the effect size of the difference score between x and y, z). The relation is:
\begin{equation}
\sigma_{z}=\sigma\sqrt{2(1-\rho)}
\end{equation}
Therefore, the relation between $d_z$ and d is $\sqrt{2(1-\rho)}$. 
As Cohen (1988) writes: "In other words, a given difference between population means for matched (dependent) samples is standardized by a value which is $\sqrt{2(1-\rho)}$ as large as would be the case were they independent". 

Cohen's $d_z$ is the effect size used in power analyses for dependent *t*-tests. 
There is no equivalent Cohen's $f_z$ for a within-participant ANOVA. 
Instead, the value for lambda ($\lambda$) is adjusted based on the correlation. 
For a one-way within-participant design lambda is calculated as in Equation \@ref(eq:lambda), multiplied by *u*, a correction for within-subject designs, calculated as:
\begin{equation}
u = \frac{k}{1-\rho}
\end{equation}
where k is the number of levels of the within-participant factor, and $\rho$ is the correlation between dependent variables.
If the correlation is 0, Equation \@ref(eq:within-n) shows that the required sample size is reduced only because each participants contributes multiple measurements, so the total number of participants required is halved.

Cohen's f is identical in within-subject and between-subject designs.
However, Equation \@ref(eq:eta-to-f) and \@ref(eq:f-to-eta) no longer hold when measurements are correlated.
The average effect size for a between design with 80 participants in each condition, means of 1, 0.5, and 0, and a standard deviation of 2 was `r power_result_1$main_results$effect_size`, while the estimated effect size increased to `r power_result_within_1$main_results$effect_size` with a correlation of 0.7 between dependent measurements.
The default settings in G\*Power expect an f or $\eta_p^2$ that does not incorporate the correlation, while the correlation is incorporated in the output of software package such as SPSS.
For a one-way within-subject design, Cohen's f can be converted into Cohen's f as SPSS provides it through:
\begin{equation}
f^2_{SPSS} = f^2 \times \frac{k}{k-1} \times \frac{n}{n-1} \times \frac{1}{1-\rho}
\end{equation}
and subsequently tranformed to $\eta_p^2$ through Equation \@ref(eq:f-to-eta).
Simulations can be used to see that because of the positive correlation between dependent variables, the effect size $\eta_p^2$ is much larger for the within-subject design (`r power_result_within_1$main_results$effect_size`) than for the between design (`r power_result_1$main_results$effect_size`). 

# Power for Interactions

The effect size for interactions in ANOVA designs depends on the pattern of means.
Let's assume the researcher plans to perform a follow-up experiment where in addition to making the voice sound cheerful or sad, a second factor is introduced by making the voice sound more robotic compared to the default human-like voice. 
Different patterns of results could be expected.
Either the same effect is observed for robotic voices, or no effect is observed for robotic voices, or the opposite effect is observed for robotic voices (a 'Marvin-the-Depressed-Robot Effect'). 
In the first case, we will only observe a main effect of voice, but in the other two scenarios there is an interaction effect between human-likeness of the voice and the emotional tone of the voice. We can start by simulating the power for a cross-over interaction for a 2x2 between-participant design with 80 participants in each group (see Figure \ref{fig:mean-plot} for the expected pattern of means). 

```{r mean-plot, fig.width=7, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE, results='hide', fig.cap="Vizualization for the expected means and standard deviations for a crossover interaction."}

design_result_cross_80 <- ANOVA_design(string = "2b*2b", n = 80, mu = c(1, 0, 0, 1), sd = 2, labelnames = c("condition", "cheerful", "sad", "voice", "human", "robot"))

power_result_cross_80 <- ANOVA_power(design_result_cross_80, alpha_level = 0.05, p_adjust = "none", nsims = nsims, seed = manuscript_seed, verbose = FALSE)

design_result_cross_40 <- ANOVA_design(string = "2b*2b", n = 40, mu = c(1, 0, 0, 1), sd = 2, labelnames = c("condition", "cheerful", "sad", "voice", "human", "robot"), plot = FALSE)

power_result_cross_40 <- ANOVA_power(design_result_cross_40, alpha_level = 0.05, p_adjust = "none", nsims = nsims, seed = manuscript_seed, verbose = FALSE)
#Analytic solution
power_twoway_between(design_result_cross_40)$power_A
power_twoway_between(design_result_cross_40)$power_B
power_twoway_between(design_result_cross_40)$power_AB
```

Mathematically the interaction effect is computed as the cell mean minus the sum of the grand mean, the marginal mean in each row minus the grand mean, and the marginal mean in each column minus grand mean. For example, for the cheerful human-like voice condition this is 1 (the value in the cell) - (0.5 [the grand mean] + 0.5 [the cell mean minus the marginal mean in row 1, 0.5] + 0.5 [the cell mean minus the marginal mean in column 2, 0.5]). 
Thus, 1 - (0.5 + (0.5) + (0.5)) = -0.5.
Completing this for all four cells gives the values -0.5, 0.5, 0.5, -0.5.
Cohen's f is then $f = \frac { \sqrt { \frac { -0.5^2 + 0.5^2 + 0.5 + -0.5^2 } { 4 } }}{ 2 } = 0.25$.
Simulations show we have `r power_result_cross_80`% power when we collected 80 participants per condition.
A cross-over (also called 'disordinal') interaction with two levels per factor has exactly the same power as the initial two-group design. if we halve the sample size per group for the cross-over interaction from 80 to 40.
Power with 40 participants per condition is `r power_result_cross_40$main_results$power[3]`%.
Main effects in an ANOVA are based on the means for one factor averaged over the other factors (e.g., the main effect of human-like versus robot-like voice, irrespective of whether it is cheerful or sad).
The interaction effect, which can be contrast coded as 1, -1, -1, 1, is similarly a test of whether the effects are non-additive based on the scores in each cell, where the null-hypothesis of no additive effect can be rejected if the deviation expected when effects in each cell would be purely additive can be rejected. 
The key insight here is that not the sample size per condition, but the total sample size dover all other factor determines the power for the main effects and the interaction [cf. @westfall_think_2015].

```{r sim-interaction-2, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
design_result_ordinal <- ANOVA_design(string = "2b*2b", n = 160, mu = c(1, 0, 0, 0), sd = 2, labelnames = c("condition", "cheerful", "sad", "voice", "human", "robot"))
power_result_ordinal <- ANOVA_power(design_result_ordinal, alpha_level = 0.05, p_adjust = "none", seed = manuscript_seed, nsims = nsims)

#Analytic solution
power_twoway_between(design_result_ordinal)$power_A
power_twoway_between(design_result_ordinal)$power_B
power_twoway_between(design_result_ordinal)$power_AB
power_twoway_between(design_result_ordinal)$Cohen_f_AB
power_twoway_between(design_result_ordinal)$eta_p_2_AB
```

We can also examine the statistical power for a pattern of results that indicated that there was no difference in interacting with a cheerful of sad conversational agent with a robot voice. 
In this case, we expect an 'ordinal' interaction (the means for the human-like voice are never lower than the means for the robot-like voice, and thus there is no cross-over effect). The pattern of means is now 1, 0, 0, 0, with only a single mean that differs from the rest.
As has been pointed out [@giner-sorolla_powering_2018; @simonsohn_no-way_2014] these designs require larger samples sizes to have the same power to detect the interaction, compared to the two-group comparison.
The reason for this is that the effect size is only half as large, with Cohen's f = 0.125 (compared to 0.25 in the cross-over interaction).
By steadily increasing the sample size in the simulation, we see that to achieve the same power as for the two-group comparison, a total sample size of 635 is required, almost four times as large as the sample size for the two-group comparison (160).

```{r sim-interaction-3, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#I am not exactly sure how this simulation is being used -- Aaron
design_result <- ANOVA_design(string = "2b*2b", n = 40, mu = c(2, 0, 0, 0), sd = 2, labelnames = c("condition", "cheerful", "sad", "voice", "human", "robot"))
#Analytic solution
power_twoway_between(design_result)$power_A
power_twoway_between(design_result)$power_B
power_twoway_between(design_result)$power_AB
power_twoway_between(design_result)$Cohen_f_AB
power_twoway_between(design_result)$eta_p_2_AB
design_result <- ANOVA_design(string = "2b*2b*2b", n = 20, mu = c(4, 0, 0, 0, 0, 0, 0, 0), sd = 2, labelnames = c("condition", "cheerful", "sad", "voice", "human", "robot", "factor_c", "c1", "c2"))
power_threeway_between(design_result)$power_ABC
power_threeway_between(design_result)$Cohen_f_ABC

```

The power in the 2x2 ordinal interaction where only one cell mean differs from the other three cell means is identical to the power we would have if the single mean was twice as far from the remaining means (for a pattern of means of 2, 0, 0, 0). 
Similarly, if we would examine a 2x2x2 interaction where only one cell differs from the other means, Cohen's f would be 0.25 when the pattern of means is 4, 0, 0, 0, 0, 0, 0, 0 across the eight cells. 
The take-home message is that a 'medium' effect size translates into a much more extreme pattern of means in an ordinal interaction than in a disordinal (crossover) interaction, or in a 2x2x2 interaction compared to a 2x2 interaction.
It might therefore be more intuitive to perform a power analysis based on the expected pattern of means, and compute Cohen's f based on this pattern, than to specify an effect size directly. 
The same effect size can represent very different patterns of means depending on the type of interaction and the number of factors (see also @perugini_practical_2018).

# Power for Simple Comparisons

Although an initial goal when performing an ANOVA might be to test the *omnibus null hypothesis*, which answers the question whether there are *any* differences between group means, we often want to know which groups differ from each other. Thus, an ANOVA is often followd up by individual comparisons (whether *planned* or *post-hoc*).
One feature of the simulation is that it provides the statistical power for all individual comparisons that can be performed.
The power and effect size estimates are based on *t*-tests. 
Taking into account the variance estimates from other groups (which the *t*-tests do not do) can have power benefits. 
However, these benefits depend on whether the homogeneity assumption is met, which is often not warranted in psychological research, and violations of the homogeneity assumption impact Type 1 error rates [@delacre_why_2018].    
Furthermore, *t*-tests allow researchers to specify directional predictions which increase the power of individual comparisons.

Power analysis for individual comparisons are relatively straightforward and can easily be done in all power analysis software.
Nevertheless, we hope that providing a complete overview of the tests a researcher might be interested in prevent researchers from performing power analyses for the ANOVA but not for independent comparisons.
Sometimes the interaction in the ANOVA will have more power than the independent comparisons (i.e., in the case of a cross-over interaction) and sometimes the power for the interaction will be lower than the power for independent comparisons.
It is always important to check if you have adequate power for all tests you plan to perform. 

# Type 1 Error Control in Exploratory ANOVA's

In a 2x2x2 design, an ANOVA will give the test results for three main effects, three two-way interactions, and one three-way interaction. 
Because seven statistical tests are performed, the probability of making at least one Type 1 error in a single exploratory 2x2x2 ANOVA is $1-(0.95)^7$ = 30%.
It is therefore important to control error rates when performing multiple comparisons [@cramer_hidden_2014].

It is possible to control the overall Type I error rate by specifying an adjustment for multiple comparisons in the simulation. 
By adjusting for multiple comparisons we ensure that we do not conclude there is an effect in any of the individual tests more often than the desired Type I error rate.
Several techniques to control error rates exist, of which the best known is the Bonferroni adjustment.
The Holm procedure is slightly more powerful than the Bonferroni adjustment, without requiring additional assumptions [for other approaches, see @bretz_multiple_2011].
Because the adjustment for multiple comparisons lowers the alpha level, it also lowers the statistical power.
When designing a study where you will perform multiple comparisons, it is important to take into account the corrected alpha level when determining the required sample size, both for the ANOVA tests, as for the independent comparisons.

```{r sim-holm, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
design_result_holm_correction <- ANOVA_design(string = "2b*2b", n = 40, mu = c(1, 0, 0, 1), sd = 2, labelnames = c("condition", "cheerful", "sad", "voice", "human", "robot"))
power_result_holm_correction <- ANOVA_power(design_result_holm_correction, alpha_level = 0.05, p_adjust = "holm", nsims = nsims, seed = manuscript_seed, verbose = FALSE)
```

If we revisit the 2x2 between-condition ANOVA where a cross-over interaction was predicted and 40 participants per condition were collected, and apply a Holm correction for multiple comparisons, we see power for the interaction is reduced from `r power_result_cross_40$main_results$power[3]`% without applying a correction, to `r power_result_holm_correction$main_results$power[3]`% after correcting for multiple comparisons.
There are three tests for a 2x2 design (two main effects, one interaction), but 6 independent comparisons between the four conditions, and thus the correction for multiple comparisons has a stronger effect on the independent comparisons.
For example, the power for the dependent *t*-test comparing two within-participant conditions falls from `r power_result_cross_40$pc_results$power[1]`% without correcting for multiple comparisons, to `r power_result_holm_correction$pc_results$power[1]`% when adjusting the alpha level for the 6 possible comparisons between the four conditions.

These simulations should reveal the cost of exploring across all possible comparisons while controlling error rates.
It is often advisable to preregister specific tests that are of interest to a researcher.
If one is really interested in all possible comparisons, but wants to control the probability of incorrectly assuming there is an effect, the reduction in power due to the lower alpha level should be counteracted by increasing the sample size to maintain an adequate level of power for both the ANOVA as the independent comparisons. 

# P-value distributions

Statistical power is the long run probability of observing a *p*-value smaller than the alpha level.
One intuitive way to illustrate this concept is to plot the distribution of *p*-values for all simulations.
The simulation code automatically stores plots for *p*-value distributions for each simulation, both for the ANOVA results, as for the independent comparisons.
In Figure \ref{fig:p-plot} we see that for the cross-over interaction with 40 participants per condition discussed earlier *p*-values are distributed uniformly for the two main effects of condition and voice. 
For the condition by voice interaction most *p*-values fall below the alpha level.
Indeed, from the `r nsims` simulations, `r power_result_cross_40$main_results$power[3]`% fall below the alpha level, which is a nice vizualization of what power means. 

```{r p-plot, fig.width=7, fig.height=10, echo=FALSE, message=FALSE, warning=FALSE, results='hide', fig.cap="Distribution of p-values for 6 independent comparisons between four conditions of a 2x2 mixed ANOVA"}
power_result_cross_40$plot1
```

# Conclusion

When designing informative studies, it is important to carefully justify the sample size.
Determining the sample size that is required for a well-powered factorial ANOVA design with multiple factors is quite complex. Existing power analysis software lacks options to calculate power for designs with multiple within-participant factors.
Simulation based approaches can help to provide insights into the factors that determine the statistical power in factorial ANOVA designs.
If researchers are able to specify the expected pattern of means, standard deviations, and correlations between variables, they can use simulations to estimate the sample size required for factorial designs that consist of multiple factors and both within- and between-subject factors
The R package and Shiny app that accompany this paper enable researchers to perform simulations for factorial experiments of up to three factors and any number of levels, making it easy to perform simulation-based power analysis without extensive programming experience.

\newpage


# References

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
