---
title: "Simulation Results"
subtitle: "Shiny App developed by Dani&euml;l Lakens and Aaron Caldwell"
output: pdf_document
date: '`r format(Sys.time(), "%B %d, %Y")`'
params:
  tablePC: NA,
  tableMain: NA,
  pvalue_plot1: NA,
  pvalue_plot2: NA,
  means_plot: NA,
  n: NA,
  padjust: NA,
  model: NA,
  design: NA,
  cor_mat: NA,
  sigmatrix: NA
  seed_number: NA
  nsims: NA
  alpha_level: NA
---




Below are the results from the ANOVA simulation app. If you encounter any problems please visit our GitHub page (https://github.com/Lakens/ANOVA_power_simulation) to the raise the issue.

**Study Design**

`r paste(params$design)`

**Model Formula** 

`r paste(params$model)`

The sample size was **`r params$n`** per cell with the following adjustment for multiple comparisons: **`r params$padjust`**. 

**Correlation Matrix**
```{r echo = FALSE}
knitr::kable(params$cor_mat)
```

**Variance-Covariance Matrix**
```{r echo = FALSE}
knitr::kable(params$sigmatrix)
```

```{r, results='asis', echo=FALSE}
cat("\\newpage")
```


```{r echo = FALSE}
params$means_plot
```


```{r, results='asis', echo=FALSE}
cat("\\newpage")
```

The seed number for this simulation was set at `r paste(params$seed_number)`. 
A total number of `r paste(params$nsims)` simulations with an alpha level of `r paste(params$alpha_level)`.
Please re-use this input to replicate the results below.

**ANOVA Power (%) and Effect Sizes (Partial Eta Squared)**

```{r echo = FALSE}
knitr::kable(params$tableMain)

params$pvalue_plot1
```

**Multiple Comparisons Power (%) and Effect Sizes (Cohen's d~z~)**

```{r echo = FALSE}
knitr::kable(params$tablePC)

params$pvalue_plot2
```



